---
title: "Data exploration: healthy human PBMCs"
author: "Belinda Phipson"
date: "31 May 2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE
)
```

# Load the libraries

```{r}
library(Seurat)
library(speckle)
library(limma)
library(ggplot2)
library(edgeR)
library(patchwork)
library(cowplot)
library(gridGraphics)
```

```{r}
set.seed(10)
```

# Read the data into R

The data is stored in a Seurat object. The cells have been classified into 
broader and more refined cell types.

```{r}
pbmc <- readRDS("./data/pool_1.rds")
```

# Visualise the data by cell type and individual 

```{r}
# Cell type information
table(pbmc$predicted.celltype.l2)
DimPlot(pbmc, group.by = "predicted.celltype.l2")
```

```{r}
table(pbmc$individual)
DimPlot(pbmc, group.by = "individual")
```

# Run the Seurat workflow for normalisation, scaling, PCA and UMAP

```{r}
pbmc <- NormalizeData(pbmc)
pbmc <- FindVariableFeatures(pbmc, selection.method = "vst", nfeatures = 2000)
pbmc <- ScaleData(pbmc)
pbmc <- RunPCA(pbmc, features = VariableFeatures(object = pbmc))
ElbowPlot(pbmc)
pbmc <- RunUMAP(pbmc, dims = 1:11)
```

# Visualise data with UMAP

```{r, fig.width=8, fig.height=5}
DimPlot(pbmc, reduction = "umap",group.by = "predicted.celltype.l1", label=TRUE, label.size=6) + theme(legend.position = "none")  + ggtitle("Broad cell type predictions")
```

```{r}
DimPlot(pbmc, reduction = "umap",group.by = "predicted.celltype.l2") + ggtitle("Refined cell type predictions")
```

```{r}
d1 <- DimPlot(pbmc, reduction = "umap",group.by = "predicted.celltype.l2") + theme(legend.position = "none") + ggtitle("a") + theme(plot.title = element_text(size = 18, hjust = 0))
```

# Explore cell type proportions among the 12 individuals

```{r, fig.width=8, fig.height=5}
props <- getTransformedProps(clusters = pbmc$predicted.celltype.l2, 
                             sample = pbmc$individual)

p1 <- plotCellTypeProps(clusters = pbmc$predicted.celltype.l2, sample = pbmc$individual) + theme(axis.text.x = element_text(angle = 45))+ ggtitle("Refined cell type proportions") + 
theme(plot.title = element_text(size = 18, hjust = 0))
p1 + theme_bw() + theme(panel.grid.major = element_blank(),
                        panel.grid.minor = element_blank()) + theme(axis.text.x = element_text(angle = 45))
```

```{r}
p2 <- plotCellTypeProps(clusters = pbmc$predicted.celltype.l1, sample = pbmc$individual)
p2 + theme(axis.text.x = element_text(angle = 45)) + ggtitle("Broad cell type proportions")
```

```{r}
pdf(file="./output/Fig1ab.pdf", width =14, height=6)
d1 + p1
dev.off()
```

# Exploring heterogeneity in cell type proportions between individuals


```{r}
counts <- table(pbmc$predicted.celltype.l2, pbmc$individual)
baselineN <- rowSums(counts)
N <- sum(baselineN)
baselineprops <- baselineN/N
```


```{r}
pbmc$final_ct <- factor(pbmc$predicted.celltype.l2, levels=names(sort(baselineprops, decreasing = TRUE)))
```

```{r}
counts <- table(pbmc$final_ct, pbmc$individual)
baselineN <- rowSums(counts)
N <- sum(baselineN)
baselineprops <- baselineN/N
```

```{r}
props <- getTransformedProps(clusters = pbmc$final_ct, 
                             sample = pbmc$individual)
```

```{r}
cols <- ggplotColors(nrow(props$Proportions))
m <- match(rownames(props$Proportions),levels(factor(pbmc$predicted.celltype.l2)))

```

```{r}
par(mfrow=c(1,1))
par(mar=c(7,5,2,2))
plot(jitter(props$Proportions[,1]), col = cols[m], pch=16, ylim=c(0,max(props$Proportions)),
     xaxt="n", xlab="", ylab="Cell type proportion", cex.lab=1.5, cex.axis=1.5)
for(i in 2:ncol(props$Proportions)){
  points(jitter(1:nrow(props$Proportions)),props$Proportions[,i], col = cols[m],
         pch=16)
}
axis(side=1, at=1:nrow(props$Proportions), las=2, 
     labels=rownames(props$Proportions))
title("Cell type proportions estimates for 12 individuals")
```

The mean-variance relationship plots below show that the data is overdispersed
compared to what would be expected under a Binomial or Poisson distribution.

```{r, fig.width=7, fig.height=6}
plotCellTypeMeanVar(counts)

plotCellTypePropsMeanVar(counts)
```

```{r,echo=FALSE, eval=FALSE}
layout(matrix(c(1,1,2,3), 2, 2, byrow = TRUE))
#layout.show(2)

par(mar=c(7,5,2,2))
plot(jitter(props$Proportions[,1]), col = cols[m], pch=16, ylim=c(0,max(props$Proportions)),
     xaxt="n", xlab="", ylab="Cell type proportion", cex.lab=1.5, cex.axis=1.5)
for(i in 2:ncol(props$Proportions)){
  points(jitter(1:nrow(props$Proportions)),props$Proportions[,i], col = cols[m],
         pch=16)
}
axis(side=1, at=1:nrow(props$Proportions), las=2, 
     labels=rownames(props$Proportions))
title("c", adj=0, cex.main=2)

plotCellTypeMeanVar(counts)
title("d", adj = 0, cex.main=2)

plotCellTypePropsMeanVar(counts)
title("e", adj = 0, cex.main=2)

```

```{r, echo=FALSE, eval=FALSE}
pdf(file="./output/Fig1cde.pdf", width =14, height=12)

layout(matrix(c(1,1,2,3), 2, 2, byrow = TRUE))
#layout.show(2)

par(mar=c(7,5,2,2))
plot(jitter(props$Proportions[,1]), col = cols[m], pch=16, ylim=c(0,max(props$Proportions)),
     xaxt="n", xlab="", ylab="Cell type proportion", cex.lab=1.5, cex.axis=1.5)
for(i in 2:ncol(props$Proportions)){
  points(jitter(1:nrow(props$Proportions)),props$Proportions[,i], col = cols[m],
         pch=16)
}
axis(side=1, at=1:nrow(props$Proportions), las=2, 
     labels=rownames(props$Proportions))
title("c", adj=0, cex.main=2)

plotCellTypeMeanVar(counts)
title("d", adj = 0, cex.main=2)

plotCellTypePropsMeanVar(counts)
title("e", adj = 0, cex.main=2)

dev.off()
```


