---
title: "Application to real single cell RNA-seq datasets"
author: "Belinda Phipson"
date: "06/03/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE
)
```

# Introduction

We analysed three different publicly available single cell datasets to highlight 
the different types of models that can be fitted within the propeller framework.

* Young and old female and male PBMCs
  + Huang, Zhaohao, Binyao Chen, Xiuxing Liu, He Li, Lihui Xie, Yuehan Gao, Runping Duan, et al. 2021. “Effects of Sex and Aging on the Immune Cell Landscape as Assessed by Single-Cell Transcriptomic Analysis.” Proceedings of the National Academy of Sciences of the United States of America 118 (33). https://doi.org/10.1073/pnas.2023216118.

* Healthy human heart biopsies across development
  + Sim, Choon Boon, Belinda Phipson, Mark Ziemann, Haloom Rafehi, Richard J. Mills, Kevin I. Watt, Kwaku D. Abu-Bonsrah, et al. 2021. “Sex-Specific Control of Human Heart Maturation by the Progesterone Receptor.” Circulation 143 (16): 1614–28.

* Bronchoalveolar lavage fluid in a COVID19 dataset
  + Liao, Mingfeng, Yang Liu, Jing Yuan, Yanling Wen, Gang Xu, Juanjuan Zhao, Lin Cheng, et al. 2020. “Single-Cell Landscape of Bronchoalveolar Immune Cells in Patients with COVID-19.” Nature Medicine 26 (6): 842–44.

For all three datasets we use the logit transformation, except for the covid dataset.
In the covid dataset, there is clearly an outlier for the plasma cell type. Using
the logit transformation, plasma is found to be significantly enriched in severe
covid, but using the arcsin square root transform shows that it is not significantly
enriched. While logit may be more powerful according to the simulations, it appears
sensitive to outliers, and in this case we would prefer to use the arcsin square root transform.

# Load libraries

```{r}
library(speckle)
library(limma)
library(edgeR)
library(pheatmap)
library(gt)
```

```{r}
source("./code/convertData.R")
```


# Young and old female and male PBMCs

This dataset was published in PNAS in 2021 and examined PBMCs from 20 individuals.
The dataset was balanced in terms of age and sex (5 samples in each group: Male + Young,
Male + Old, Female + Young, Female + Old).

From the supplementary data I have extracted the cell type proportions and using 
information from the paper I have converted the proportions into the necessary
data type (a list object) to analyse with propeller.


```{r}
sexCT <- read.delim("./data/CTpropsTransposed.txt", row.names = 1)
sexprops <- sexCT/100
prop.list <- convertDataToList(sexprops,data.type="proportions", transform="logit",
                               scale.fac=174684/20)

sampleinfo <- read.csv("./data/sampleinfo.csv", nrows = 20)
celltypes <- read.csv("./data/CelltypeLevels.csv")
group.immune <- paste(sampleinfo$Sex, sampleinfo$Age, sep=".")
```

```{r}
gt(data.frame(table(sampleinfo$Sex, sampleinfo$Age)), rownames_to_stub = TRUE, caption="Sample info for aging dataset")
```

Set up design matrix using a means model, taking into account age and sex.

```{r}
designAS <- model.matrix(~0+sampleinfo$Age + sampleinfo$Sex)
colnames(designAS) <- c("old","young","MvsF")

# Young vs old
mycontr <- makeContrasts(young-old, levels=designAS)
propeller.ttest(prop.list = prop.list,design = designAS, contrasts = mycontr,
                robust=TRUE,trend=FALSE,sort=TRUE)

```

Visualise significant cell types:
```{r, fig.width=11, fig.height=6}
par(mfrow=c(1,2))
stripchart(as.numeric(sexprops["CD8.Naive",])~group.immune,
           vertical=TRUE, pch=c(8,16), method="jitter",
           col = c(ggplotColors(20)[20],"hotpink",4, "darkblue"),cex=2,
           ylab="Proportions", cex.axis=1.25, cex.lab=1.5,
           group.names=c("F.Old","F.Young","M.Old","M.Young"))
title("CD8.Naive: Young Vs Old", cex.main=1.5, adj=0)
text(3.2,0.18, labels = "Adj.Pval = 0.004")

stripchart(as.numeric(sexprops["CD16",])~group.immune,
           vertical=TRUE, pch=c(8,16), method="jitter",
           col = c(ggplotColors(20)[20],"hotpink",4, "darkblue"),cex=2,
           ylab="Proportions", cex.axis=1.25, cex.lab=1.5,
           group.names=c("F.Old","F.Young","M.Old","M.Young"))
title("CD16: Young Vs Old", cex.main=1.5, adj=0)
text(2.2,0.049, labels = "Adj.Pval = 0.023")
```


# Heart development analysis

This dataset was published by Sim et al in 2021 and looked at heart development
across fetal, young and adult developmental time points. There are three samples
at each developmental time point. There is also mix of male 
and female samples and one of the key findings of the study was transcriptional
differences in cardiomyocyte development between males and females.

Here we look at differences in cellular composition of human hearts across the
developmental trajectory, taking sex into account.

We can simply perform an anova test to find any differences between the three 
time points. The propeller.anova function can be called to do this directly.

We can also look at development as a continuous trajectory and model development
as a continuous variable by getting the transformed proportions and using fitting
functions in limma directly.

Both of these analysis approaches are shown below.

```{r}
heart.info <- read.csv(file="./data/cellinfo.csv", row.names = 1)
heart.counts <- table(heart.info$Celltype, heart.info$Sample)
trueprops <- rowSums(heart.counts)/sum(rowSums(heart.counts))

heart.info$Group <- NA
heart.info$Group[grep("f",heart.info$Sample)] <- "fetal"
heart.info$Group[grep("y",heart.info$Sample)] <- "young"
heart.info$Group[grep("a",heart.info$Sample)] <- "adult"

sample <- factor(heart.info$Sample, levels= paste(rep(c("f","y","a"), each=3),c(1:3),sep=""))
group <- factor(heart.info$Group, levels=c("fetal","young","adult"))
```

```{r}
grp <- factor(rep(c("fetal","young","adult"),each=3), levels=c("fetal","young","adult"))
sex <- factor(c("m","m","f","m","f","m","f","m","m"))
dose <- rep(c(1,2,3), each=3) 
```

The sample information is summarised below:

```{r}
gt(data.frame(Sample=1:9,Group=grp, Sex=sex),caption="Sample info for heart dataset")
```


## Anova test with propeller.anova

```{r}
prop.logit <- getTransformedProps(clusters = heart.info$Celltype, sample=sample,
                                  transform = "logit")
design.anova <- model.matrix(~0+grp+sex)

propeller.anova(prop.logit,design = design.anova, coef=c(1,2,3), robust=TRUE, 
                trend = FALSE, sort=TRUE)
```


## Modelling development as a continuous variable

```{r}
des.dose <- model.matrix(~dose + sex)
des.dose

fit <- lmFit(prop.logit$TransformedProps,des.dose)
fit <- eBayes(fit, robust=TRUE)
topTable(fit,coef=2)

fit.plot <- lmFit(prop.logit$Proportions,des.dose)
fit.plot <- eBayes(fit.plot, robust=TRUE)
```

The two analysis identify the same cell types as significantly enriched/depleted, 
although there is a change in the order of significance.

Three significant cell types are visualised below.

```{r,fig.width=12, fig.height=5}
par(mfrow=c(1,3))
stripchart(as.numeric(prop.logit$Proportions["Immune cells",])~grp,
           vertical=TRUE, pch=16, method="jitter",
           col = ggplotColors(4),cex=2, 
           ylab="Proportions",cex.axis=1.25, cex.lab=1.5)
title("Immune development", cex.main=1.5, adj=0)
abline(a=fit.plot$coefficients["Immune cells",1], b=fit.plot$coefficients["Immune cells",2], lty=2, lwd=2)


stripchart(as.numeric(prop.logit$Proportions["Cardiomyocytes",])~grp,
           vertical=TRUE, pch=16, method="jitter",
           col = ggplotColors(4),cex=2, 
           ylab="Proportions",cex.axis=1.25, cex.lab=1.5)
title("Cardiomyocyte development", cex.main=1.5, adj=0)
abline(a=fit.plot$coefficients["Cardiomyocytes",1], b=fit.plot$coefficients["Cardiomyocytes",2], lty=2, lwd=2)
text(2.6,0.77, labels = "Adj.Pval = 0.01")

stripchart(as.numeric(prop.logit$Proportions["Fibroblast",])~grp,
           vertical=TRUE, pch=16, method="jitter",
           col = ggplotColors(4),cex=2, 
           ylab="Proportions",cex.axis=1.25, cex.lab=1.5)
title("Fibroblast development", cex.main=1.5, adj=0)
abline(a=fit.plot$coefficients["Fibroblast",1], b=fit.plot$coefficients["Fibroblast",2], lty=2, lwd=2)
```

# COVID data

This dataset was published by Liao et al in 2020 in Nature Medicine. They compared 
moderate and severe covid to healthy controls. They sampled bronchoalveolar lavage 
fluid from each individual.

For this dataset there are no additional covariates and we use the propeller()
function with the cell level annotation information for cell type, sample, and 
group. The function automatically detects more than two groups and performs
an ANOVA for each cell type. We fit with both the logit and arcsin square root
transformed data to show the effect of an outlier sample in the plasma cell type.

```{r}
covid <- read.delim("./data/covid.cell.annotation.meta.txt")
```

propeller using logit transformed proportions:

```{r}
output.logit <- propeller(clusters=covid$celltype, sample=covid$sample_new, group=covid$group, transform="logit")
output.logit
```

propeller using arcsin square root transformed proportions:

```{r}
output.asin <- propeller(clusters=covid$celltype, sample=covid$sample_new, group=covid$group, transform="asin")
output.asin
```

```{r}
props.covid <- getTransformedProps(clusters=covid$celltype, sample=covid$sample_new,
                                   transform="logit")
```


```{r}
par(mfrow=c(1,1))
grp.covid <- rep(c("Control","Moderate","Severe"), c(4, 3, 6))
stripchart(as.numeric(props.covid$Proportions["Neutrophil",])~grp.covid,
           vertical=TRUE, pch=16, method="jitter",
           col = c(4,"purple",2),cex=2, 
           ylab="Proportions",cex.axis=1.25, cex.lab=1.5)
title("Neutrophils in severe covid", cex.main=1.5, adj=0)
text(1.5,0.195, labels = "Adj.Pval = 3.5e-06")
```

All significant cell types
```{r,fig.width=10, fig.height=10}
par(mfrow=c(2,2))
stripchart(as.numeric(props.covid$Proportions["Neutrophil",])~grp.covid,
           vertical=TRUE, pch=16, method="jitter",
           col = c(4,"purple",2),cex=2, 
           ylab="Proportions",cex.axis=1.25, cex.lab=1.5)
title("Neutrophils in covid", cex.main=1.5, adj=0)

stripchart(as.numeric(props.covid$Proportions["Plasma",])~grp.covid,
           vertical=TRUE, pch=16, method="jitter",
           col = c(4,"purple",2),cex=2, 
           ylab="Proportions",cex.axis=1.25, cex.lab=1.5)
title("Plasma in covid - outlier sample", cex.main=1.5, adj=0)

stripchart(as.numeric(props.covid$Proportions["pDC",])~grp.covid,
           vertical=TRUE, pch=16, method="jitter",
           col = c(4,"purple",2),cex=2, 
           ylab="Proportions",cex.axis=1.25, cex.lab=1.5)
title("pDC in covid", cex.main=1.5, adj=0)

stripchart(as.numeric(props.covid$Proportions["NK",])~grp.covid,
           vertical=TRUE, pch=16, method="jitter",
           col = c(4,"purple",2),cex=2, 
           ylab="Proportions",cex.axis=1.25, cex.lab=1.5)
title("NK in covid", cex.main=1.5, adj=0)
```



