---
title: "Simulations with true cell type proportion differences"
author: "Belinda Phipson"
date: "01/06/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE
)
```

# Load the libraries

```{r}
library(speckle)
library(limma)
library(edgeR)
library(pheatmap)
```

Source the simulation code:
```{r}
source("./code/SimCodeTrueDiff.R")
source("./code/auroc.R")
```

# Hierarchical model for simulating cell type proportions

I am simulating cell type proportions in a hierarchical manner. 

* The total number of cells, $n_j$, for each sample $j$, are drawn from a negative binomial
distribution with mean 5000 and dispersion 20. 
* The true cell type proportions for 7 cell types are estimated from a human heart dataset
* The sample proportion $p_{ij}$ for cell type $i$ and sample $j$ is assumed to be drawn 
from a Beta distribution with parameters $\alpha$ and $\beta$. 
* The count for cell type $i$ and sample $j$ is then drawn from a binomial
distribution with probability $p_{ij}$ and size $n_j$.

The Beta-Binomial model allows for biological variability to be simulated between samples.
The paramaters of the Beta distribution, $\alpha$ and $\beta$, determine how variable the
$p_{ij}$ will be. Larger values of $\alpha$ and $\beta$ result in a more precise 
distribution centred around the true proportions, while smaller values result in a more diffuse prior. 
The parameters for the Beta distribution were estimated from the cell type counts observed
in the human heart dataset using the function estimateBetaParamsFromCounts in the
speckle package.

# Two group simulation set up

I will generate cell type counts for 7 cell types, assuming two experimental
groups with a sample size of n=(3,5,10,20) in each group. I will calculate 
p-values from the following models:

* propeller (arcsin sqrt transformation)
* propeller (logit transformation)
* chi-square test of differences in proportions
* beta-binomial model using alternative parameterisation in edgeR
* logistic binomial regression (beta-binomial with dispersion=0)
* negative binomial regression (LRT and QLF in edgeR)
* Poisson regression (negative binomial with dispersion=0)
* CODA model

One thousand simulation datasets will be generated. 

First I set up the simulation parameters and set up the objects to capture the
output.


```{r}
# Sim parameters
set.seed(10)
nsim <- 1000
depth <- 5000

# True cell type proportions from human heart dataset
heart.info <- read.csv(file="./data/cellinfo.csv", row.names = 1)
heart.counts <- table(heart.info$Celltype, heart.info$Sample)
heart.counts <- heart.counts[-4,]
trueprops <- rowSums(heart.counts)/sum(rowSums(heart.counts))

```

```{r}
betaparams <- estimateBetaParamsFromCounts(heart.counts)

# Parameters for beta distribution
a <- betaparams$alpha
b <- betaparams$beta

# Decide on what output to keep
pval.chsq <- pval.bb <- pval.lb <- pval.nb <- pval.qlf <- pval.pois <- pval.logit <-  pval.asin <- 
  pval.coda <- matrix(NA,nrow=length(trueprops),ncol=nsim)
```

Set up true proportions for the two groups:
```{r}
# Set up true props for the two groups
grp1.trueprops <- grp2.trueprops <- trueprops
grp2.trueprops[1] <- grp1.trueprops[1]/2
grp2.trueprops[4] <- grp2.trueprops[4]*2
grp2.trueprops[7] <- grp1.trueprops[7]*3

grp2.trueprops[1] <- grp2.trueprops[1] + (1-sum(grp2.trueprops))/2
grp2.trueprops[4] <- grp2.trueprops[4] + (1-sum(grp2.trueprops))
 
sum(grp1.trueprops)
sum(grp2.trueprops)
da.fac <- grp2.trueprops/grp1.trueprops
```

```{r}
o <- order(trueprops)
par(mar=c(9,5,2,2))
barplot(t(cbind(grp1.trueprops[o],grp2.trueprops[o])), beside=TRUE, col=c(2,4), 
        las=2, ylab="True cell type proportion")
legend("topleft", fill=c(2,4),legend=c("Group 1","Group 2"))
title("True cell type proportions for Group 1 and 2")
```

```{r}
# Get hyperparameters for alpha and beta
# Note group 1 and group 2 have different b parameters to accommodate true
# differences in cell type proportions
a <- a
b.grp1 <- a*(1-grp1.trueprops)/grp1.trueprops
b.grp2 <- a*(1-grp2.trueprops)/grp2.trueprops
```

Next we simulate the cell type counts and run the various statistical models
for testing cell type proportion differences between the two groups. We expect 
to see significant differences in cell type proportions in three cell types, 
and no significant differences in the remaining four cell types between group 1 
and group 2. We expect differences in the Smooth muscle cells (most rare), 
Fibroblasts (second most abundant) and Cardiomyocytes (most abundant).

## Sample size of 3 in each group 

```{r}
nsamp <- 6

for(i in 1:nsim){
    #Simulate cell type counts
    counts <- SimulateCellCountsTrueDiff(props=trueprops,nsamp=nsamp,depth=depth,a=a,
                                         b.grp1=b.grp1,b.grp2=b.grp2)

    tot.cells <- colSums(counts)
    # propeller
    est.props <- t(t(counts)/tot.cells)
    
    #asin transform
    trans.prop <- asin(sqrt(est.props))
    
    #logit transform
    nc <- normCounts(counts)
    est.props.logit <- t(t(nc+0.5)/(colSums(nc+0.5)))
    logit.prop <- log(est.props.logit/(1-est.props.logit))

    grp <- rep(c(0,1), each=nsamp/2)
    des <- model.matrix(~grp)
  
    # asinsqrt transform
    fit <- lmFit(trans.prop, des)
    fit <- eBayes(fit, robust=TRUE)

    pval.asin[,i] <- fit$p.value[,2]
    
    # logit transform
    fit.logit <- lmFit(logit.prop, des)
    fit.logit <- eBayes(fit.logit, robust=TRUE)

    pval.logit[,i] <- fit.logit$p.value[,2]

    # Chi-square test for differences in proportions
    n <- tapply(tot.cells, grp, sum)
    for(h in 1:nrow(counts)){
        pval.chsq[h,i] <- prop.test(tapply(counts[h,],grp,sum),n)$p.value
    }

    # Beta binomial implemented in edgeR (methylation workflow)
    meth.counts <- counts
    unmeth.counts <- t(tot.cells - t(counts))
    new.counts <- cbind(meth.counts,unmeth.counts)
    sam.info <- data.frame(Sample = rep(1:nsamp,2), Group=rep(grp,2), Meth = rep(c("me","un"), each=nsamp))

    design.samples <- model.matrix(~0+factor(sam.info$Sample))
    colnames(design.samples) <- paste("S",1:nsamp,sep="")
    design.group <- model.matrix(~0+factor(sam.info$Group))   
    colnames(design.group) <- c("A","B")
    design.bb <- cbind(design.samples, (sam.info$Meth=="me") * design.group)
    lib.size = rep(tot.cells,2)

    y <- DGEList(new.counts)
    y$samples$lib.size <- lib.size
    y <- estimateDisp(y, design.bb, trend="none")
    fit.bb <- glmFit(y, design.bb)
    contr <- makeContrasts(Grp=B-A, levels=design.bb)
    lrt <- glmLRT(fit.bb, contrast=contr)
    pval.bb[,i] <- lrt$table$PValue

    # Logistic binomial regression
    fit.lb <- glmFit(y, design.bb, dispersion = 0)
    lrt.lb <- glmLRT(fit.lb, contrast=contr)
    pval.lb[,i] <- lrt.lb$table$PValue

    # Negative binomial
    y.nb <- DGEList(counts)
    y.nb <- estimateDisp(y.nb, des, trend="none")
    fit.nb <- glmFit(y.nb, des)
    lrt.nb <- glmLRT(fit.nb, coef=2)
    pval.nb[,i] <- lrt.nb$table$PValue
    
    # Negative binomial QLF test
    fit.qlf <- glmQLFit(y.nb, des, robust=TRUE, abundance.trend = FALSE)
    res.qlf <- glmQLFTest(fit.qlf, coef=2)
    pval.qlf[,i] <- res.qlf$table$PValue

    # Poisson
    fit.poi <- glmFit(y.nb, des, dispersion = 0)
    lrt.poi <- glmLRT(fit.poi, coef=2)
    pval.pois[,i] <- lrt.poi$table$PValue
    
    # CODA
    # Replace zero counts with 0.5 so that the geometric mean always works
    if(any(counts==0)) counts[counts==0] <- 0.5
    geomean <- apply(counts,2, function(x) exp(mean(log(x))))
    geomean.mat <- expandAsMatrix(geomean,dim=c(nrow(counts),ncol(counts)),byrow = FALSE)
    clr <- counts/geomean.mat
    logratio <- log(clr)
    
    fit.coda <- lmFit(logratio, des)
    fit.coda <- eBayes(fit.coda, robust=TRUE)

    pval.coda[,i] <- fit.coda$p.value[,2]

}

```


We can look at the number of significant tests at certain p-value cut-offs:
```{r}
pcut <- 0.05
de <- da.fac != 1
sig.disc <- matrix(NA,nrow=length(trueprops),ncol=9)
rownames(sig.disc) <- names(trueprops)
colnames(sig.disc) <- c("chisq","logbin","pois","asin", "logit","betabin","negbin","nbQLF","CODA")

sig.disc[,1]<-rowSums(pval.chsq<pcut)/nsim 
sig.disc[,2]<-rowSums(pval.lb<pcut)/nsim
sig.disc[,3]<-rowSums(pval.pois<pcut)/nsim 
sig.disc[,4]<-rowSums(pval.asin<pcut)/nsim 
sig.disc[,5]<-rowSums(pval.logit<pcut)/nsim 
sig.disc[,6]<-rowSums(pval.bb<pcut)/nsim
sig.disc[,7]<-rowSums(pval.nb<pcut)/nsim 
sig.disc[,8]<-rowSums(pval.qlf<pcut)/nsim
sig.disc[,9]<-rowSums(pval.coda<pcut)/nsim 
o <- order(trueprops)
sig.disc[o,]
```


```{r}
layout(matrix(c(1,1,1,2), 1, 4, byrow = TRUE))
par(mar=c(8,5,3,2))
par(mgp=c(3, 0.5, 0))
o <- order(trueprops)
names <- c("propeller (asin)","propeller (logit)","betabin","negbin","negbinQLF","CODA")
barplot(sig.disc[o,4:9],beside=TRUE,col=ggplotColors(length(b)),
        ylab="Proportion sig. tests", names=names,
        cex.axis = 1.5, cex.lab=1.5, cex.names = 1.35, ylim=c(0,1), las=2)
title(paste("Significant tests, n=",nsamp/2,sep=""), cex.main=1.5)
abline(h=pcut,lty=2,lwd=2)
par(mar=c(0,0,0,0))
plot(1, type = "n", xlab = "", ylab = "", xaxt="n",yaxt="n", bty="n")
legend("center", legend=paste("True p =",round(trueprops,3)[o]), fill=ggplotColors(length(b)), cex=1.5)


```


```{r}
o <- order(trueprops)
mysig <- sig.disc[o,4:9]
colnames(mysig) <- names
pheatmap(mysig, scale="none", cluster_rows = FALSE, cluster_cols = FALSE,
         labels_row = c(expression(paste(pi, " = 0.008*", sep="")),
                        expression(paste(pi, " = 0.016", sep="")),
                        expression(paste(pi, " = 0.064", sep="")),
                        expression(paste(pi, " = 0.076", sep="")),
                        expression(paste(pi, " = 0.102", sep="")),
                        expression(paste(pi, " = 0.183*", sep="")),
                        expression(paste(pi, " = 0.551*", sep=""))),
         main=paste("Significant tests, n=",nsamp/2,sep=""))

```


```{r}
auc.asin <- auc.logit <- auc.bb <- auc.nb <- auc.qlf <- auc.coda <- rep(NA,nsim)
for(i in 1:nsim){
  auc.asin[i] <- auroc(score=1-pval.asin[,i],bool=de)
  auc.logit[i] <- auroc(score=1-pval.logit[,i],bool=de)
  auc.bb[i] <- auroc(score=1-pval.bb[,i],bool=de)
  auc.nb[i] <- auroc(score=1-pval.nb[,i],bool=de)
  auc.qlf[i] <- auroc(score=1-pval.qlf[,i],bool=de)
  auc.coda[i] <- auroc(score=1-pval.coda[,i],bool=de)
}

mean(auc.asin)
mean(auc.logit)
mean(auc.bb)
mean(auc.nb)
mean(auc.qlf)
mean(auc.coda)
```

```{r}
par(mfrow=c(1,1))
par(mar=c(9,5,3,2))
barplot(c(mean(auc.asin),mean(auc.logit),mean(auc.bb),mean(auc.nb),mean(auc.qlf),mean(auc.coda)), ylim=c(0,1), ylab= "AUC", cex.axis=1.5, cex.lab=1.5, names=names, las=2, cex.names = 1.5)
title(paste("AUC: sample size n=",nsamp/2,sep=""),cex.main=1.5)
```

```{r}
sig.disc3 <- sig.disc
```

## Sample size of 5 in each group 

```{r}
nsamp <- 10

for(i in 1:nsim){
    #Simulate cell type counts
    counts <- SimulateCellCountsTrueDiff(props=trueprops,nsamp=nsamp,depth=depth,a=a,
                                         b.grp1=b.grp1,b.grp2=b.grp2)

    tot.cells <- colSums(counts)
    # propeller
    est.props <- t(t(counts)/tot.cells)
    
    #asin transform
    trans.prop <- asin(sqrt(est.props))
    
    #logit transform
    nc <- normCounts(counts)
    est.props.logit <- t(t(nc+0.5)/(colSums(nc+0.5)))
    logit.prop <- log(est.props.logit/(1-est.props.logit))

    grp <- rep(c(0,1), each=nsamp/2)
    des <- model.matrix(~grp)
  
    # asinsqrt transform
    fit <- lmFit(trans.prop, des)
    fit <- eBayes(fit, robust=TRUE)

    pval.asin[,i] <- fit$p.value[,2]
    
    # logit transform
    fit.logit <- lmFit(logit.prop, des)
    fit.logit <- eBayes(fit.logit, robust=TRUE)

    pval.logit[,i] <- fit.logit$p.value[,2]

    # Chi-square test for differences in proportions
    n <- tapply(tot.cells, grp, sum)
    for(h in 1:nrow(counts)){
        pval.chsq[h,i] <- prop.test(tapply(counts[h,],grp,sum),n)$p.value
    }

    # Beta binomial implemented in edgeR (methylation workflow)
    meth.counts <- counts
    unmeth.counts <- t(tot.cells - t(counts))
    new.counts <- cbind(meth.counts,unmeth.counts)
    sam.info <- data.frame(Sample = rep(1:nsamp,2), Group=rep(grp,2), Meth = rep(c("me","un"), each=nsamp))

    design.samples <- model.matrix(~0+factor(sam.info$Sample))
    colnames(design.samples) <- paste("S",1:nsamp,sep="")
    design.group <- model.matrix(~0+factor(sam.info$Group))   
    colnames(design.group) <- c("A","B")
    design.bb <- cbind(design.samples, (sam.info$Meth=="me") * design.group)
    lib.size = rep(tot.cells,2)

    y <- DGEList(new.counts)
    y$samples$lib.size <- lib.size
    y <- estimateDisp(y, design.bb, trend="none")
    fit.bb <- glmFit(y, design.bb)
    contr <- makeContrasts(Grp=B-A, levels=design.bb)
    lrt <- glmLRT(fit.bb, contrast=contr)
    pval.bb[,i] <- lrt$table$PValue

    # Logistic binomial regression
    fit.lb <- glmFit(y, design.bb, dispersion = 0)
    lrt.lb <- glmLRT(fit.lb, contrast=contr)
    pval.lb[,i] <- lrt.lb$table$PValue

    # Negative binomial
    y.nb <- DGEList(counts)
    y.nb <- estimateDisp(y.nb, des, trend="none")
    fit.nb <- glmFit(y.nb, des)
    lrt.nb <- glmLRT(fit.nb, coef=2)
    pval.nb[,i] <- lrt.nb$table$PValue
    
    # Negative binomial QLF test
    fit.qlf <- glmQLFit(y.nb, des, robust=TRUE, abundance.trend = FALSE)
    res.qlf <- glmQLFTest(fit.qlf, coef=2)
    pval.qlf[,i] <- res.qlf$table$PValue

    # Poisson
    fit.poi <- glmFit(y.nb, des, dispersion = 0)
    lrt.poi <- glmLRT(fit.poi, coef=2)
    pval.pois[,i] <- lrt.poi$table$PValue
    
    # CODA
    # Replace zero counts with 0.5 so that the geometric mean always works
    if(any(counts==0)) counts[counts==0] <- 0.5
    geomean <- apply(counts,2, function(x) exp(mean(log(x))))
    geomean.mat <- expandAsMatrix(geomean,dim=c(nrow(counts),ncol(counts)),byrow = FALSE)
    clr <- counts/geomean.mat
    logratio <- log(clr)
    
    fit.coda <- lmFit(logratio, des)
    fit.coda <- eBayes(fit.coda, robust=TRUE)

    pval.coda[,i] <- fit.coda$p.value[,2]

}

```


We can look at the number of significant tests at certain p-value cut-offs:
```{r}
pcut <- 0.05
de <- da.fac != 1
sig.disc <- matrix(NA,nrow=length(trueprops),ncol=9)
rownames(sig.disc) <- names(trueprops)
colnames(sig.disc) <- c("chisq","logbin","pois","asin", "logit","betabin","negbin","nbQLF","CODA")

sig.disc[,1]<-rowSums(pval.chsq<pcut)/nsim 
sig.disc[,2]<-rowSums(pval.lb<pcut)/nsim
sig.disc[,3]<-rowSums(pval.pois<pcut)/nsim 
sig.disc[,4]<-rowSums(pval.asin<pcut)/nsim 
sig.disc[,5]<-rowSums(pval.logit<pcut)/nsim 
sig.disc[,6]<-rowSums(pval.bb<pcut)/nsim
sig.disc[,7]<-rowSums(pval.nb<pcut)/nsim 
sig.disc[,8]<-rowSums(pval.qlf<pcut)/nsim
sig.disc[,9]<-rowSums(pval.coda<pcut)/nsim 
o <- order(trueprops)
sig.disc[o,]
```


```{r}
layout(matrix(c(1,1,1,2), 1, 4, byrow = TRUE))
par(mar=c(8,5,3,2))
par(mgp=c(3, 0.5, 0))
o <- order(trueprops)
names <- c("propeller (asin)","propeller (logit)","betabin","negbin","negbinQLF","CODA")
barplot(sig.disc[o,4:9],beside=TRUE,col=ggplotColors(length(b)),
        ylab="Proportion sig. tests", names=names,
        cex.axis = 1.5, cex.lab=1.5, cex.names = 1.35, ylim=c(0,1), las=2)
title(paste("Significant tests, n=",nsamp/2,sep=""), cex.main=1.5)
abline(h=pcut,lty=2,lwd=2)
par(mar=c(0,0,0,0))
plot(1, type = "n", xlab = "", ylab = "", xaxt="n",yaxt="n", bty="n")
legend("center", legend=paste("True p =",round(trueprops,3)[o]), fill=ggplotColors(length(b)), cex=1.5)


```


```{r}
o <- order(trueprops)
mysig <- sig.disc[o,4:9]
colnames(mysig) <- names
pheatmap(mysig, scale="none", cluster_rows = FALSE, cluster_cols = FALSE,
         labels_row = c(expression(paste(pi, " = 0.008*", sep="")),
                        expression(paste(pi, " = 0.016", sep="")),
                        expression(paste(pi, " = 0.064", sep="")),
                        expression(paste(pi, " = 0.076", sep="")),
                        expression(paste(pi, " = 0.102", sep="")),
                        expression(paste(pi, " = 0.183*", sep="")),
                        expression(paste(pi, " = 0.551*", sep=""))),
         main=paste("Significant tests, n=",nsamp/2,sep=""))

```


```{r}
auc.asin <- auc.logit <- auc.bb <- auc.nb <- auc.qlf <- auc.coda <- rep(NA,nsim)
for(i in 1:nsim){
  auc.asin[i] <- auroc(score=1-pval.asin[,i],bool=de)
  auc.logit[i] <- auroc(score=1-pval.logit[,i],bool=de)
  auc.bb[i] <- auroc(score=1-pval.bb[,i],bool=de)
  auc.nb[i] <- auroc(score=1-pval.nb[,i],bool=de)
  auc.qlf[i] <- auroc(score=1-pval.qlf[,i],bool=de)
  auc.coda[i] <- auroc(score=1-pval.coda[,i],bool=de)
}

mean(auc.asin)
mean(auc.logit)
mean(auc.bb)
mean(auc.nb)
mean(auc.qlf)
mean(auc.coda)
```

```{r}
par(mfrow=c(1,1))
par(mar=c(9,5,3,2))
barplot(c(mean(auc.asin),mean(auc.logit),mean(auc.bb),mean(auc.nb),mean(auc.qlf),mean(auc.coda)), ylim=c(0,1), ylab= "AUC", cex.axis=1.5, cex.lab=1.5, names=names, las=2, cex.names = 1.5)
title(paste("AUC: sample size n=",nsamp/2,sep=""),cex.main=1.5)
```

```{r}
sig.disc5 <- sig.disc
```


## Sample size of 10 in each group 

```{r}
nsamp <- 20

for(i in 1:nsim){
    #Simulate cell type counts
    counts <- SimulateCellCountsTrueDiff(props=trueprops,nsamp=nsamp,depth=depth,a=a,
                                         b.grp1=b.grp1,b.grp2=b.grp2)

    tot.cells <- colSums(counts)
    # propeller
    est.props <- t(t(counts)/tot.cells)
    
    #asin transform
    trans.prop <- asin(sqrt(est.props))
    
    #logit transform
    nc <- normCounts(counts)
    est.props.logit <- t(t(nc+0.5)/(colSums(nc+0.5)))
    logit.prop <- log(est.props.logit/(1-est.props.logit))

    grp <- rep(c(0,1), each=nsamp/2)
    des <- model.matrix(~grp)
  
    # asinsqrt transform
    fit <- lmFit(trans.prop, des)
    fit <- eBayes(fit, robust=TRUE)

    pval.asin[,i] <- fit$p.value[,2]
    
    # logit transform
    fit.logit <- lmFit(logit.prop, des)
    fit.logit <- eBayes(fit.logit, robust=TRUE)

    pval.logit[,i] <- fit.logit$p.value[,2]

    # Chi-square test for differences in proportions
    n <- tapply(tot.cells, grp, sum)
    for(h in 1:nrow(counts)){
        pval.chsq[h,i] <- prop.test(tapply(counts[h,],grp,sum),n)$p.value
    }

    # Beta binomial implemented in edgeR (methylation workflow)
    meth.counts <- counts
    unmeth.counts <- t(tot.cells - t(counts))
    new.counts <- cbind(meth.counts,unmeth.counts)
    sam.info <- data.frame(Sample = rep(1:nsamp,2), Group=rep(grp,2), Meth = rep(c("me","un"), each=nsamp))

    design.samples <- model.matrix(~0+factor(sam.info$Sample))
    colnames(design.samples) <- paste("S",1:nsamp,sep="")
    design.group <- model.matrix(~0+factor(sam.info$Group))   
    colnames(design.group) <- c("A","B")
    design.bb <- cbind(design.samples, (sam.info$Meth=="me") * design.group)
    lib.size = rep(tot.cells,2)

    y <- DGEList(new.counts)
    y$samples$lib.size <- lib.size
    y <- estimateDisp(y, design.bb, trend="none")
    fit.bb <- glmFit(y, design.bb)
    contr <- makeContrasts(Grp=B-A, levels=design.bb)
    lrt <- glmLRT(fit.bb, contrast=contr)
    pval.bb[,i] <- lrt$table$PValue

    # Logistic binomial regression
    fit.lb <- glmFit(y, design.bb, dispersion = 0)
    lrt.lb <- glmLRT(fit.lb, contrast=contr)
    pval.lb[,i] <- lrt.lb$table$PValue

    # Negative binomial
    y.nb <- DGEList(counts)
    y.nb <- estimateDisp(y.nb, des, trend="none")
    fit.nb <- glmFit(y.nb, des)
    lrt.nb <- glmLRT(fit.nb, coef=2)
    pval.nb[,i] <- lrt.nb$table$PValue
    
    # Negative binomial QLF test
    fit.qlf <- glmQLFit(y.nb, des, robust=TRUE, abundance.trend = FALSE)
    res.qlf <- glmQLFTest(fit.qlf, coef=2)
    pval.qlf[,i] <- res.qlf$table$PValue

    # Poisson
    fit.poi <- glmFit(y.nb, des, dispersion = 0)
    lrt.poi <- glmLRT(fit.poi, coef=2)
    pval.pois[,i] <- lrt.poi$table$PValue
    
    # CODA
    # Replace zero counts with 0.5 so that the geometric mean always works
    if(any(counts==0)) counts[counts==0] <- 0.5
    geomean <- apply(counts,2, function(x) exp(mean(log(x))))
    geomean.mat <- expandAsMatrix(geomean,dim=c(nrow(counts),ncol(counts)),byrow = FALSE)
    clr <- counts/geomean.mat
    logratio <- log(clr)
    
    fit.coda <- lmFit(logratio, des)
    fit.coda <- eBayes(fit.coda, robust=TRUE)

    pval.coda[,i] <- fit.coda$p.value[,2]

}

```


We can look at the number of significant tests at certain p-value cut-offs:
```{r}
pcut <- 0.05
de <- da.fac != 1
sig.disc <- matrix(NA,nrow=length(trueprops),ncol=9)
rownames(sig.disc) <- names(trueprops)
colnames(sig.disc) <- c("chisq","logbin","pois","asin", "logit","betabin","negbin","nbQLF","CODA")

sig.disc[,1]<-rowSums(pval.chsq<pcut)/nsim 
sig.disc[,2]<-rowSums(pval.lb<pcut)/nsim
sig.disc[,3]<-rowSums(pval.pois<pcut)/nsim 
sig.disc[,4]<-rowSums(pval.asin<pcut)/nsim 
sig.disc[,5]<-rowSums(pval.logit<pcut)/nsim 
sig.disc[,6]<-rowSums(pval.bb<pcut)/nsim
sig.disc[,7]<-rowSums(pval.nb<pcut)/nsim 
sig.disc[,8]<-rowSums(pval.qlf<pcut)/nsim
sig.disc[,9]<-rowSums(pval.coda<pcut)/nsim 
o <- order(trueprops)
sig.disc[o,]
```


```{r}
layout(matrix(c(1,1,1,2), 1, 4, byrow = TRUE))
par(mar=c(8,5,3,2))
par(mgp=c(3, 0.5, 0))
o <- order(trueprops)
names <- c("propeller (asin)","propeller (logit)","betabin","negbin","negbinQLF","CODA")
barplot(sig.disc[o,4:9],beside=TRUE,col=ggplotColors(length(b)),
        ylab="Proportion sig. tests", names=names,
        cex.axis = 1.5, cex.lab=1.5, cex.names = 1.35, ylim=c(0,1), las=2)
title(paste("Significant tests, n=",nsamp/2,sep=""), cex.main=1.5)
abline(h=pcut,lty=2,lwd=2)
par(mar=c(0,0,0,0))
plot(1, type = "n", xlab = "", ylab = "", xaxt="n",yaxt="n", bty="n")
legend("center", legend=paste("True p =",round(trueprops,3)[o]), fill=ggplotColors(length(b)), cex=1.5)


```


```{r}
o <- order(trueprops)
mysig <- sig.disc[o,4:9]
colnames(mysig) <- names
pheatmap(mysig, scale="none", cluster_rows = FALSE, cluster_cols = FALSE,
         labels_row = c(expression(paste(pi, " = 0.008*", sep="")),
                        expression(paste(pi, " = 0.016", sep="")),
                        expression(paste(pi, " = 0.064", sep="")),
                        expression(paste(pi, " = 0.076", sep="")),
                        expression(paste(pi, " = 0.102", sep="")),
                        expression(paste(pi, " = 0.183*", sep="")),
                        expression(paste(pi, " = 0.551*", sep=""))),
         main=paste("Significant tests, n=",nsamp/2,sep=""))

```


```{r}
auc.asin <- auc.logit <- auc.bb <- auc.nb <- auc.qlf <- auc.coda <- rep(NA,nsim)
for(i in 1:nsim){
  auc.asin[i] <- auroc(score=1-pval.asin[,i],bool=de)
  auc.logit[i] <- auroc(score=1-pval.logit[,i],bool=de)
  auc.bb[i] <- auroc(score=1-pval.bb[,i],bool=de)
  auc.nb[i] <- auroc(score=1-pval.nb[,i],bool=de)
  auc.qlf[i] <- auroc(score=1-pval.qlf[,i],bool=de)
  auc.coda[i] <- auroc(score=1-pval.coda[,i],bool=de)
}

mean(auc.asin)
mean(auc.logit)
mean(auc.bb)
mean(auc.nb)
mean(auc.qlf)
mean(auc.coda)
```

```{r}
par(mfrow=c(1,1))
par(mar=c(9,5,3,2))
barplot(c(mean(auc.asin),mean(auc.logit),mean(auc.bb),mean(auc.nb),mean(auc.qlf),mean(auc.coda)), ylim=c(0,1), ylab= "AUC", cex.axis=1.5, cex.lab=1.5, names=names, las=2, cex.names = 1.5)
title(paste("AUC: sample size n=",nsamp/2,sep=""),cex.main=1.5)
```

```{r}
sig.disc10 <- sig.disc
```

## Sample size of 20 in each group 

```{r}
nsamp <- 40

for(i in 1:nsim){
    #Simulate cell type counts
    counts <- SimulateCellCountsTrueDiff(props=trueprops,nsamp=nsamp,depth=depth,a=a,
                                         b.grp1=b.grp1,b.grp2=b.grp2)

    tot.cells <- colSums(counts)
    # propeller
    est.props <- t(t(counts)/tot.cells)
    
    #asin transform
    trans.prop <- asin(sqrt(est.props))
    
    #logit transform
    nc <- normCounts(counts)
    est.props.logit <- t(t(nc+0.5)/(colSums(nc+0.5)))
    logit.prop <- log(est.props.logit/(1-est.props.logit))

    grp <- rep(c(0,1), each=nsamp/2)
    des <- model.matrix(~grp)
  
    # asinsqrt transform
    fit <- lmFit(trans.prop, des)
    fit <- eBayes(fit, robust=TRUE)

    pval.asin[,i] <- fit$p.value[,2]
    
    # logit transform
    fit.logit <- lmFit(logit.prop, des)
    fit.logit <- eBayes(fit.logit, robust=TRUE)

    pval.logit[,i] <- fit.logit$p.value[,2]

    # Chi-square test for differences in proportions
    n <- tapply(tot.cells, grp, sum)
    for(h in 1:nrow(counts)){
        pval.chsq[h,i] <- prop.test(tapply(counts[h,],grp,sum),n)$p.value
    }

    # Beta binomial implemented in edgeR (methylation workflow)
    meth.counts <- counts
    unmeth.counts <- t(tot.cells - t(counts))
    new.counts <- cbind(meth.counts,unmeth.counts)
    sam.info <- data.frame(Sample = rep(1:nsamp,2), Group=rep(grp,2), Meth = rep(c("me","un"), each=nsamp))

    design.samples <- model.matrix(~0+factor(sam.info$Sample))
    colnames(design.samples) <- paste("S",1:nsamp,sep="")
    design.group <- model.matrix(~0+factor(sam.info$Group))   
    colnames(design.group) <- c("A","B")
    design.bb <- cbind(design.samples, (sam.info$Meth=="me") * design.group)
    lib.size = rep(tot.cells,2)

    y <- DGEList(new.counts)
    y$samples$lib.size <- lib.size
    y <- estimateDisp(y, design.bb, trend="none")
    fit.bb <- glmFit(y, design.bb)
    contr <- makeContrasts(Grp=B-A, levels=design.bb)
    lrt <- glmLRT(fit.bb, contrast=contr)
    pval.bb[,i] <- lrt$table$PValue

    # Logistic binomial regression
    fit.lb <- glmFit(y, design.bb, dispersion = 0)
    lrt.lb <- glmLRT(fit.lb, contrast=contr)
    pval.lb[,i] <- lrt.lb$table$PValue

    # Negative binomial
    y.nb <- DGEList(counts)
    y.nb <- estimateDisp(y.nb, des, trend="none")
    fit.nb <- glmFit(y.nb, des)
    lrt.nb <- glmLRT(fit.nb, coef=2)
    pval.nb[,i] <- lrt.nb$table$PValue
    
    # Negative binomial QLF test
    fit.qlf <- glmQLFit(y.nb, des, robust=TRUE, abundance.trend = FALSE)
    res.qlf <- glmQLFTest(fit.qlf, coef=2)
    pval.qlf[,i] <- res.qlf$table$PValue

    # Poisson
    fit.poi <- glmFit(y.nb, des, dispersion = 0)
    lrt.poi <- glmLRT(fit.poi, coef=2)
    pval.pois[,i] <- lrt.poi$table$PValue
    
    # CODA
    # Replace zero counts with 0.5 so that the geometric mean always works
    if(any(counts==0)) counts[counts==0] <- 0.5
    geomean <- apply(counts,2, function(x) exp(mean(log(x))))
    geomean.mat <- expandAsMatrix(geomean,dim=c(nrow(counts),ncol(counts)),byrow = FALSE)
    clr <- counts/geomean.mat
    logratio <- log(clr)
    
    fit.coda <- lmFit(logratio, des)
    fit.coda <- eBayes(fit.coda, robust=TRUE)

    pval.coda[,i] <- fit.coda$p.value[,2]

}

```


We can look at the number of significant tests at certain p-value cut-offs:
```{r}
pcut <- 0.05
de <- da.fac != 1
sig.disc <- matrix(NA,nrow=length(trueprops),ncol=9)
rownames(sig.disc) <- names(trueprops)
colnames(sig.disc) <- c("chisq","logbin","pois","asin", "logit","betabin","negbin","nbQLF","CODA")

sig.disc[,1]<-rowSums(pval.chsq<pcut)/nsim 
sig.disc[,2]<-rowSums(pval.lb<pcut)/nsim
sig.disc[,3]<-rowSums(pval.pois<pcut)/nsim 
sig.disc[,4]<-rowSums(pval.asin<pcut)/nsim 
sig.disc[,5]<-rowSums(pval.logit<pcut)/nsim 
sig.disc[,6]<-rowSums(pval.bb<pcut)/nsim
sig.disc[,7]<-rowSums(pval.nb<pcut)/nsim 
sig.disc[,8]<-rowSums(pval.qlf<pcut)/nsim
sig.disc[,9]<-rowSums(pval.coda<pcut)/nsim 
o <- order(trueprops)
sig.disc[o,]
```


```{r}
layout(matrix(c(1,1,1,2), 1, 4, byrow = TRUE))
par(mar=c(8,5,3,2))
par(mgp=c(3, 0.5, 0))
o <- order(trueprops)
names <- c("propeller (asin)","propeller (logit)","betabin","negbin","negbinQLF","CODA")
barplot(sig.disc[o,4:9],beside=TRUE,col=ggplotColors(length(b)),
        ylab="Proportion sig. tests", names=names,
        cex.axis = 1.5, cex.lab=1.5, cex.names = 1.35, ylim=c(0,1), las=2)
title(paste("Significant tests, n=",nsamp/2,sep=""), cex.main=1.5)
abline(h=pcut,lty=2,lwd=2)
par(mar=c(0,0,0,0))
plot(1, type = "n", xlab = "", ylab = "", xaxt="n",yaxt="n", bty="n")
legend("center", legend=paste("True p =",round(trueprops,3)[o]), fill=ggplotColors(length(b)), cex=1.5)


```


```{r}
o <- order(trueprops)
mysig <- sig.disc[o,4:9]
colnames(mysig) <- names
pheatmap(mysig, scale="none", cluster_rows = FALSE, cluster_cols = FALSE,
         labels_row = c(expression(paste(pi, " = 0.008*", sep="")),
                        expression(paste(pi, " = 0.016", sep="")),
                        expression(paste(pi, " = 0.064", sep="")),
                        expression(paste(pi, " = 0.076", sep="")),
                        expression(paste(pi, " = 0.102", sep="")),
                        expression(paste(pi, " = 0.183*", sep="")),
                        expression(paste(pi, " = 0.551*", sep=""))),
         main=paste("Significant tests, n=",nsamp/2,sep=""))

```



```{r}
sig.disc20 <- sig.disc
```


```{r}
auc.asin <- auc.logit <- auc.bb <- auc.nb <- auc.qlf <- auc.coda <- rep(NA,nsim)
for(i in 1:nsim){
  auc.asin[i] <- auroc(score=1-pval.asin[,i],bool=de)
  auc.logit[i] <- auroc(score=1-pval.logit[,i],bool=de)
  auc.bb[i] <- auroc(score=1-pval.bb[,i],bool=de)
  auc.nb[i] <- auroc(score=1-pval.nb[,i],bool=de)
  auc.qlf[i] <- auroc(score=1-pval.qlf[,i],bool=de)
  auc.coda[i] <- auroc(score=1-pval.coda[,i],bool=de)
}

mean(auc.asin)
mean(auc.logit)
mean(auc.bb)
mean(auc.nb)
mean(auc.qlf)
mean(auc.coda)
```

```{r}
par(mfrow=c(1,1))
par(mar=c(9,5,3,2))
barplot(c(mean(auc.asin),mean(auc.logit),mean(auc.bb),mean(auc.nb),mean(auc.qlf),mean(auc.coda)), ylim=c(0,1), ylab= "AUC", cex.axis=1.5, cex.lab=1.5, names=names, las=2, cex.names = 1.5)
title(paste("AUC: sample size n=",nsamp/2,sep=""),cex.main=1.5)
```

